---
title: "Scatterplots"
format: live-html
engine: knitr
webr:
  packages: ['ggformula', 'palmerpenguins']
execute: 
  warning: false
  message: false
code-copy: true
code-overflow: wrap
toc: true
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

## Getting Started

First, be sure you have installed `ggformula`. Remember, you only need to install the package once on your machine.

Then, be sure to load the package `ggformula`. Remember, you need to do this with each new Quarto/RMarkdown document or R Session.

```{webr}
#| label: setup
library(ggformula) #for graphs
```

### Data for Examples

As a reminder (see [Overview of Data Visualization](visualization-overview.qmd)), we will be using the `penguins` data from the `palmerpenguins` package:

```{r}
#| label: load-penguins
library(palmerpenguins)
```

Here is a snippet of the data:

```{r}
#| label: data-structure 
#| echo: false
library(dplyr)
library(gt)
slice_sample(penguins, n = 5) |> 
  gt::gt() |> 
  gt::tab_header(title = "Palmer Penguins")
```

## Scatterplots with Two Quantitative Variables

### Basic Code

To visualize the relationship between two **quantitative** variables `x` and `y`, here is the general structure for a scatterplot.

```{r}
#| label: scatter-structure
#| eval: false
gf_point(y ~ x, 
         data = mydata)
```

The `y` position will map to the y-axis, and the `x` position will map to the x-axis. 


Run the code below to see an example using the **quantitative** variables `bill_length_mm` and `bill_depth_mm` from the `penguins` data. Then flip the placement of the variables in the `y` and `x` positions. Then replace `bill_length_mm` with another quantitative variable from the `penguins` data (e.g. `flipper_length_mm`)

```{webr}
#| label: scatter-penguins
gf_point(bill_depth_mm ~ bill_length_mm, 
         data = penguins) 
```

**Notice** the warning produced from running the code. This is just a warning that there were rows (penguins) ignored due to missing data for the variables visualized.



### Adding Labels

Descriptive labels are important for any visualization. We can always add them to any visualization by adding `xlab =` and `ylab =` to your function. 

```{r}
#| label: one-var-structure-labels
#| eval: false
gf_point(y ~ x, 
         data = mydata,
         xlab = "X Axis Label",
         ylab = "Y Axis Label",
         title = "Descriptive Title") 
```

Add labels and a title to the histogram for `bill_length_mm`.

```{webr}
#| label: scatter-add-labels
gf_point(bill_depth_mm ~ bill_length_mm, 
         data = penguins)
```

### Other Modifications

We can add a few other modifications that purely aesthetic - just to make our graphs look nicer or easier to read.

#### Adding the Data Points to the Graph

Boxplots often hide the amount of data they represent (the number of cases). To combat this, we can add the data points to the boxplot using the `gf_jitter()` function.  You can modify the `height = ` argument to change the distance between points. Notice the `0` which centers the points on the y-axis at 0. 

```{r}
#| label: one-var-structure-jitter
#| eval: false
gf_boxplot(~x, 
           data = mydata) |> 
  gf_jitter(0 ~ x, 
           data = mydata,
           height = 1) |> 
  gf_theme(axis.ticks.y = element_blank(),  
           axis.text.y = element_blank())   
```


Modify the values for `height = ` agrument and see how they affect the point distribution on the y-axis. Remember, the y-axis has no meaning here, so the heights of the points are random and not representative of any element of the data. Just the position on the x-axis is meaningful. 

```{webr}
#| label: boxplot-add-jitter
gf_boxplot(~bill_length_mm, 
           data = penguins) |> 
  gf_jitter(0 ~ bill_length_mm, 
           data = penguins,
           height = 0.25) |> 
  gf_theme(axis.ticks.y = element_blank(), 
           axis.text.y = element_blank())
```


#### Filling the Boxes with Color

We can add a color to fill the boxes by telling R to fill the bars with a specified color either using a [built in color from R](https://r-charts.com/colors/) or using a [hex code for colors](https://htmlcolorcodes.com/color-picker/) .

```{r}
#| label: one-var-structure-box-fill
#| eval: false
gf_boxplot(~x, 
           data = mydata,
           xlab = "X Axis Label",
           title = "Descriptive Title",
           fill = "darkcyan") |> 
  gf_theme(axis.ticks.y = element_blank(), 
           axis.text.y = element_blank())
```

#### Changing the Theme

The package `ggformula` is built on top of another package called `ggplot2` and so any `ggplot2` function can be added to a `ggformula` generated graphic. For example, we can change the theme to a [built-in theme](https://ggplot2.tidyverse.org/reference/ggtheme.html). To connect `ggplot2` functions, we will use a `+` at the end of the line precedding the function.  

Try changing the theme to the following graph:

```{webr}
#| label: boxplot-one-theme
gf_boxplot(~bill_length_mm, 
           data = penguins) |> 
  gf_theme(axis.ticks.y = element_blank(), 
           axis.text.y = element_blank()) + #notice the plus +
  theme_light() #ggplot2 function
```


## Boxplots for Comparisons Across Groups

When we have a **quantitative** variable that has been measured across multiple groups, we may be interested in comparing boxplots across the values/groups of a **categorical** variable. We can do this by adding a y-axis variable that represents the *categorical* variable.  

```{r}
#| label: two-var-structure
#| eval: false
gf_boxplot(y ~ x, 
           data = mydata) 
#we don't need to do anything to the y-axis now because it represents a variable here 
```

The variable places in the `y` position will be on the y-axis and the variable in the `x` position will be on the x-axis. So we can change the orientation of our boxplot by switching the position of the **quantitative** and **categorical** variable.  

Here is an example with `species` on the y-axis. Try modifying the code so that `species` is on the x-axis and `bill_length_mm` is on the y-axis. Then modify the axes labels.
```{webr}
#| label: two-var-boxplot
gf_boxplot(species ~ bill_length_mm, 
           data = penguins,
           xlab = "Add Variable X Information",
           ylab = "Add Variable Y Information") 
```

### Adding Jitter to Groups

It is much easier to add jitter points to the boxplot across groups, we just pipe into the `gf_jitter()` function.  Modify the `height =` argument to adjust the random position of the points on the y-axis.

```{webr}
#| label: two-var-boxplot-jitter
gf_boxplot(species ~ bill_length_mm, 
           data = penguins,
           xlab = "Add Variable X Information",
           ylab = "Add Variable Y Information") |> 
  gf_jitter(height = 0.5)
```


### Adding Color to Groups  

Similar to changing the color of boxes to a single color, we can use the `fill = ` argument associated with the **categorical variable**. 

Here is the boxplot of `bill_length_mm` with fill color varied by `species` a categorical variable with values of Adelie, Chinstrap, and Gentoo.  Modify the code below to change the fill color to another categorical variable such as `island` or `sex` and see what happens.  

```{webr}
#| label: two-var-boxplot-color
gf_boxplot(species ~ bill_length_mm, 
           fill = ~species,
           data = penguins,
           xlab = "Add Variable X Information",
           ylab = "Add Variable Y Information",
           show.legend = FALSE) #hides unnecessary legend for fill
```


If we want to specify color and add jitter, we can do that too!

```{webr}
#| label: two-var-boxplot-color-jitter
gf_boxplot(species ~ bill_length_mm, 
           fill = ~species,
           data = penguins,
           xlab = "Add Variable X Information",
           ylab = "Add Variable Y Information",
           show.legend = FALSE,
           alpha = 0.5) |> #makes the fill more transparent 
  gf_jitter(color = ~species,
            height = 0.25,
            show.legend = FALSE) #hides new legend for color

```