---
title: "Generating Multiple Summary Statistics"
format: 
  live-html:
    webr:
      packages: 
        - mosaic
        - ggformula
        - palmerpenguins
      cell-options:
        warning: false
        completion: false
execute: 
  warning: false
code-copy: true
code-overflow: wrap
toc: true
css: ../styles.css
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ../_extensions/r-wasm/live/_gradethis.qmd >}}


```{webr}
#| label: theme
#| include: false
#| edit: false
library(ggplot2)
theme_set(theme_light())
options(ggplot2.continuous.color = scale_color_viridis_c, 
        ggplot2.continuous.fill = scale_fill_viridis_c,
        ggplot2.discrete.color = scale_color_viridis_d,
        ggplot2.discrete.fill = scale_fill_viridis_d)
```


## Getting Started

First, be sure you have installed `ggformula`. Remember, you only need to install each package once on your machine.

Then, be sure to load the packages `ggformula`. Remember, you need to do this with each new Quarto/RMarkdown document or R Session.

```{webr}
#| label: setup
library(ggformula) #for summary stats
```


### Data for Examples

As a reminder (see [Overview of Data ation](summary-stats-overview.qmd)), we will be using the `penguins` data from the `palmerpenguins` package:

```{r}
#| label: load-penguins
library(palmerpenguins)
```

Here is a snippet of the data:

```{r}
#| label: data-structure 
#| echo: false
library(dplyr) #for slice_sample function
library(gt) #for table formatting
slice_sample(penguins, n = 5) |> 
  gt::gt() |> 
  gt::tab_header(title = "Palmer Penguins")
```


### Multiple Statistics for a Single Variable

For a single quantitative variable, `x`, here is the general structure for calculating a multiple statistics in R using the `df_stats()` function from the `ggformula` package.

```{r}
#| label: one-var-structure-stats
#| eval: false
df_stats(~x, data = mydata)
```

Run the code below to see an example using the **quantitative** variable `bill_length_mm` from the `penguins` data. Then replace `bill_length_mm` with another quantitative variable from the `penguins` data (e.g. `bill_depth_mm`)

```{webr}
#| label: stats-one
df_stats(~bill_length_mm, data = penguins)
```



### Multiple Statistics Across Groups

When we want to calculate the multiple summary statistics of a **quantitative** variable (`y`) measured across the values/groups of a **categorical** variable (`x`)

```{r}
#| label: two-var-structure-stats
#| eval: false
df_stats(y ~ x, data = mydata)
```

Run the code below to see an example using the **quantitative** variable `bill_length_mm` and `species` from the `penguins` data. Then replace `bill_length_mm` with another quantitative variable from the `penguins` data (e.g. `bill_depth_mm`) and/or another categorical variable (e.g., `sex`). 


```{webr}
#| label: stats-two
df_stats(bill_length_mm ~ species, data = penguins)
```



