---
title: "Measures of Center"
format: 
  live-html:
    webr:
      packages: 
        - mosaic
        - palmerpenguins
      cell-options:
        warning: false
        completion: false
execute: 
  warning: false
code-copy: true
code-overflow: wrap
toc: true
css: ../styles.css
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}


```{webr}
#| label: theme
#| include: false
#| edit: false
library(ggplot2)
theme_set(theme_light())
options(ggplot2.continuous.color = scale_color_viridis_c, 
        ggplot2.continuous.fill = scale_fill_viridis_c,
        ggplot2.discrete.color = scale_color_viridis_d,
        ggplot2.discrete.fill = scale_fill_viridis_d)
```


## Getting Started

First, be sure you have installed `mosaic`. Remember, you only need to install each package once on your machine.

Then, be sure to load the packages `mosaic`. Remember, you need to do this with each new Quarto/RMarkdown document or R Session.

```{webr}
#| label: setup
library(mosaic) #for summary stats
```


### Data for Examples <a class="btn btn-primary" data-bs-toggle="collapse" href="#data-for-examples" role="button" aria-expanded="false" aria-controls="data-for-examples">Show</a>

::: {#data-for-examples .collapse .multi-collapse}

As a reminder (see [Overview of Summary Statistics](summary-stats-overview.qmd)), we will be using the `penguins` data from the `palmerpenguins` package:

```{r}
#| label: load-penguins
library(palmerpenguins)
```

Here is a snippet of the data:

```{r}
#| label: data-structure 
#| echo: false
library(dplyr) #for slice_sample function
library(gt) #for table formatting
slice_sample(penguins, n = 5) |> 
  gt::gt() |> 
  gt::tab_header(title = "Palmer Penguins")
```

:::

## Mean

::: {.callout-note collapse="true"}

## Definition: MEAN
Sometimes called "the average" it is a summary of a **quantitative** variable observed within a sample. It measures the "fulcrum" or the balancing point of the data values and uses every value within its calculation.  

$$\bar{x} = \frac{x_1 + x_2 + x_3 + ... + x_n}{n} = \frac{1}{n}\sum_{i = 1}^{n}{x_i} $$

:::


### Basic Code

For a single **quantitative** variable, `x`, here is the general structure for calculating a mean in R using the `mean()` function from the `mosaic` package.

```{r}
#| label: one-var-structure-mean
#| eval: false
mean(~x, data = mydata)
```

Run the code below to see an example using the **quantitative** variable `bill_length_mm` from the `penguins` data. Then replace `bill_length_mm` with another quantitative variable from the `penguins` data (e.g. `bill_depth_mm`).

```{webr}
#| label: mean-one
mean(~bill_length_mm, data = penguins)
```

::: {.callout-warning}
## Handling Missing Values

Notice the returned value of `NA`. The function needs to have another argument added that tells R to ignore missing values (`NA`) in order to calculate the mean, `na.rm = TRUE`. Add the argument to the code above and rerun it. The warning will disappear and a numeric value (the mean) will be provided.

```{r}
#| label: one-var-structure-na-mean
#| eval: false
mean(~x, data = mydata, na.rm = TRUE)
```

:::


### Multiple Means Across Groups

When we want to calculate the mean of a **quantitative** variable (`y`) measured across the values/groups of a **categorical** variable (`x`) it is a simple modification to the single mean code.

```{r}
#| label: two-var-structure-mean
#| eval: false
mean(y ~ x, data = mydata)
```

Run the code below to see an example using the **quantitative** variable `bill_length_mm` and `species` from the `penguins` data. Then replace `bill_length_mm` with another quantitative variable from the `penguins` data (e.g. `bill_depth_mm`) and/or another categorical variable (e.g., `sex`). 

Remember to add `na.rm = TRUE` if you get a *warning*. 

```{webr}
#| label: mean-two
mean(bill_length_mm ~ species, data = penguins)
```


## Median

:::{.callout-note collapse="true"}

## Definition: MEDIAN
Also known as the 50th percentile and the 2nd quantile, the median is the value within the data such that roughly 50% of the other data points have a value greater than it and the other 50% of other data points have a value less than it. It is the middle value of the data.

:::


### Basic Code

For a single quantitative variable, `x`, here is the general structure for calculating a median in R using the `median()` function from the `mosaic` package.

```{r}
#| label: one-var-structure-median
#| eval: false
median(~x, data = mydata)
```

Run the code below to see an example using the **quantitative** variable `bill_length_mm` from the `penguins` data. Then replace `bill_length_mm` with another quantitative variable from the `penguins` data (e.g. `bill_depth_mm`)

```{webr}
#| label: median-one
median(~bill_length_mm, data = penguins)
```

::: {.callout-warning collapse="true"}
## Handling Missing Values
Notice the returned value of `NA`. The function needs to have another argument added that tells R to ignore missing values (`NA`) in order to calculate the median, `na.rm = TRUE`. Add the argument to the code above

```{r}
#| label: one-var-structure-na-median
#| eval: false
median(~x, data = mydata, na.rm = TRUE)
```

:::


### Multiple Medians Across Groups

When we want to calculate the median of a **quantitative** variable (`y`) measured across the values/groups of a **categorical** variable (`x`) it is a simple modification to the single median code.

```{r}
#| label: two-var-structure-median
#| eval: false
median(y ~ x, data = mydata)
```

Run the code below to see an example using the **quantitative** variable `bill_length_mm` and `species` from the `penguins` data. Then replace `bill_length_mm` with another quantitative variable from the `penguins` data (e.g. `bill_depth_mm`) and/or another categorical variable (e.g., `sex`). 

Remember to add `na.rm = TRUE` if you get a *warning*. 

```{webr}
#| label: median-two
median(bill_length_mm ~ species, data = penguins)
```


